Topological Equivariant Artist Model

### 2021-04-06 paper/paper.tex
post passover housekeeping

%% declare the category of your paper, only shown in review mode
\vgtccategory{Research}
%% please declare the paper type of your paper to help reviewers, only shown in review mode
%% choices:
%% * algorithm/technique
%% * application/design study
%% * evaluation
%% * system
%% * theory/model
\vgtcpapertype{Theory/Model}

@@

General purpose visualization libraries are often tasked with supporting many domains, leading the libraries to grow organically in ways that can lead to incoherent interfaces and brittle components. To adapt to modern library development needs, this paper presents a functional model of the how components provided by lower level libraries can be guaranteed to be structure preserving maps from data to graphic. We target this level of library architecture because our aim is the libraries developers use to build automated and user facing tools, not those tools themselves; our model is agnostic as to what makes for a good visualization so long as the structure preserving constraints are satisfied. 
\begin{figure}[H]
  \centering 
  \subfloat[\label{fig:intro:continuity}]{%
    \includegraphics[width=1\columnwidth]{continuity.png}}\\
  \subfloat[\label{fig:intro:equivariance}]{%
    \includegraphics[width=1\columnwidth]{equivariance.png}}
\caption{In \autoref{fig:intro:continuity}, the line plot does not preserve \textit{continuity} because it implies that the discrete records are connected to each other, while the bar plot is \textit{continuity} preserving because it visually represents the records as independent data points. In \autoref{fig:intro:equivariance}, the data in blue is scaled by a factor of two, yielding the data in orange. To preserve \textit{equivariance}, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
\label{fig:intro:structure}
\end{figure}
The structure that components of a visualization library must preserve is \textit{continuity} and \textit{equivariance}, as illustrated in \autoref{fig:intro:structure}. \textit{Continuity} is the way in which the records in the dataset are connected to each other and \textit{equivariance} means that if any action is applied to the data or the visual, for example a rotation, permutation, translation, or rescaling, an equivalent action is applied on the other side. In \autoref{fig:intro:continuity}, the line plot does not preserve continuity because the line connecting the discrete categories implies that the frequency of weather events is sampled from a continuous interval and the categories are points on that interval. In \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from orange to blue.

To specify the constraints components must satisfy to be equivariant continuity preserving mappings from data to visual representation, we developed a functional model of library architecture, which we call the Topological Equivariant Artist Model (TEAM). Motivated by the challenge of re-architecturing the Python visualization library Matplotlib, a functional architecture derived from our model would gain a generalized data abstraction and well specified reusable components with minimal side effects\cite{loudenProgrammingLanguagesPrinciples2010}, in turn yielding improvements in maintainability, extendability, scaling, and support for concurrency. The contribution of this work is 
\begin{enumerate}
  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:math:graphic:base}
  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:math:artist}
  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:code}
\end{enumerate}

\section{Related Work}
The notion that visualization is equivariant continuity preserving maps from data to visual representation is neither a new formalism nor a new implementation goal, but this work bridges the formalism and implementation in a functional manner with a topological approach at a building blocks library\cite{wongsuphasawatNavigatingWideWorld2021} level. When introducing the retinal variables, Bertin informally specifies that continuity is preserved in the mark and defines equivariance constraints in terms of data and visual variables being selective, associative, ordered, or quantitative\cite{bertinSemiologyGraphicsDiagrams2011a}. In the \textit{A Presentation Tool}(APT) model, Mackinlay embeds the continuity constraint in the choice of visualization type and generalizes the equivariance constraint to preserving a binary operator from one domain to another. The algebraic model of visualization\cite{kindlmannAlgebraicProcessVisualization2014}, proposed by Kindlmann and Scheidegger, restricts equivariance to invertible transformations. Our model defines equivariance in terms of monoid actions, which are more restrictive than all binary operations but more general than invertible transformations. Their algebraic model also describes visualization as commutative mappings from data space to representation space to graphic space. Our model has the same notion of commutative mappings across analogous spaces, but uses topology to explicitly include continuity.

We emphasize continuity because visualization tools tend to either be architectured around a core data structure\cite{HeerSoftware2006} or explicitly implement the visual algorithm in terms of the data continuity the algorithm expects \cite{toryRethinkingVisualizationHighlevel2004}. For example, the relational database is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}. Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}. Neither the table nor image nor graph model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.  Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can be rather inconsistent as every visualization has a different notion of how the data is structured. Our model facilitates unifying these APIs via an abstraction of data general enough to encompass most continuities, the clear separation of data, visual, and graphic transformations into functional components to mitigate side effects, and by specifying the constraints these components must satisfy to map data to visualizations in a structure preserving manner.

@@

In \autoref{fig:teaser}, we illustrate the output of a minimal Q that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and an image. The scatter plot can be defined as 
\begin{equation}
\vmark(xpos, ypos)(\alpha, \beta)
\label{eq:math:artist:q:scatter}
\end{equation}
with a constant \textit{size} and color $\gsection_{RGB} = (0,0,0)$ that are defined as part of \vmark. The position of this swatch of color can be computed relative to the location on the disc \((\alpha, \beta) \in \gbase_{\dbasepoint}\) as shown in \autoref{fig:math:artist:base}
\begin{align*}
x &= size *\alpha \cos(\beta) + xpos \\
y &= size *\alpha \sin(\beta) + ypos
\end{align*}
such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black. 
In contrast, the line plot 
\begin{equation}
\vmark(xpos, \hat{n}_{1}, ypos, \hat{n}_{2})(\alpha, \beta)
\label{eq:math:artist:q:line}
\end{equation}
in \autoref{fig:teaser} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase. As shown in \autoref{fig:math:artist:base}, line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness; therefore the artist takes as input the jet bundle \cite{JetBundle2020,musilovaCalculusVariationsJet2016} $\mathcal{J}^{2}(\dtotal)$ which is the data \dtotal\ and the first and second derivatives of \dtotal. The magnitude of the slope is $\lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2}$
such that the normal is  $\hat{n}_{1} = \frac{n_1}{\lvert n \rvert}, \; \hat{n}_{2} = \frac{n_2}{\lvert n \rvert}$ which yields components of \gsection\
\begin{align*}
 x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n}_1(\vindex(\alpha)) \\
 y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n}_2(\vindex(\alpha)) 
\end{align*}
where (x,y) look up the position $\vindex(\alpha)$ on the data and the derivatives $\hat{n}_1, \hat{n}_2$ . The derivatives are then multiplied by a $width$ parameter to specify the thickness.
In \autoref{fig:teaser}, the image 
\begin{equation}
\vmark(xpos, ypos, color)
\label{eq:math:artist:q:image}
\end{equation}
 is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$. The indexing variables $(\alpha, \beta)$ define the distance along the space, which is then used by \vindex\ to map into \dbase\ to lookup the color values 
\begin{equation*}
R = R(\vindex(\alpha, \beta)), \; G = G(\vindex(\alpha, \beta)), \; B = B(\vindex(\alpha, \beta))
\end{equation*}
In the case of an image, the indexing mapper \vindex\ may do some translating to a convention expected by \vmark, for example reorientng the array such that the first row in the data is at the bottom of the graphic.

@@

\subsubsection{Visual Encoders}
The visual parameter serves as the dictionary key because the visual representation is constructed from the encoding applied to the data  $\vsection = \vchannel \circ \dsection$. For the scatter plot, the mappings for the visual fiber components $\vfiber=(x,y, facecolors, s)$ are defined as
\begin{minted}{python}
cmap =  color.Categorical({'true':'deeppink', 
                           'false':'deepskyblue'}) 
# {P_i name:{'name':c_i, 'encoder:\nu_i}}
V = {'x': {'name': 'v4', 'encoder': lambda x: x}, 
     'y': {'name': 'v2', 'encoder': lambda x: x},
     'facecolors': {'name':'v3', 'encoder': cmap}, 
     's':{'name': None , 
          'encoder': lambda _: itertools.repeat(.02)}}
\end{minted}
where \mintinline{python}{lambda x: x} is an identity \vchannel, \mintinline{python}|{'name':None}| maps into \vfiber\ without corresponding \dsection\ to set a constant visual value, and \mintinline{python}{color.Categorical} is a custom \vchannel\ implemented as a class for reusability.  A test for equivariance, as described in \autoref{eq:math:artist:nu:equivariance}, can be implemented trivially
\begin{minted}{python}
#\nu_i(m_r(E_i)) = \varphi(m_r)(\nu_i((E_i))
def test_nominal(values, encoder):
    m1 = list(zip(values, encoder(values)))
    random.shuffle(values)
    m2 = list(zip(values, encoder(values)))
    assert sorted(m1) == sorted(m2)
\end{minted}
but is currently factored out of the artist for clarity.

@@

\section{Discussion}
This work contributes a functional model of the structure-preserving maps from data to visual representation to guide the development of visualization libraries, thereby providing a means to express the constraints of preserving the data continuity in the graphic and faithfully translating the properties of the data variables into visual variables. Combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets this model support a variety of datasets, including  discrete relational tables, multivariate high resolution spatio temporal datasets, and complex networks. Decomposing the artist into encoding \vchannel, assembly \vmark, and reindexing \vindex\ provides the specifications that the graphic must have continuity equivalent to the data, and that the visual characteristics of the graphics are equivariant to their corresponding components under monoid actions. This model defines these constraints on the transformation function such that they are not specific to any one type of encoding or visual characteristic. Encoding the graphic space as a fiber bundle provides a structure rich abstraction of the target graphical design in the target display space. The toy prototype built using this model validates that is usable for a general purpose visualization tool since it can be iteratively integrated into the existing architecture rather than starting from scratch. Factoring out graphic formation into assembly functions allows for much more clarity in how they differ. This prototype demonstrates that this framework can generate the fundemental point (scatter plot) and line (line chart) marks.

@@

\subsection{Future Work}
More work is needed to formalize the composition operators and equivalence class \vartisteq and  we need to implement artists that demonstrate that the model can underpin a minimally viable library, foremost an image\cite{haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as a boxplot\cite{wickham40YearsBoxplots2011}. Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014}. While this paper formulates visualization in terms of monoidal action homomorphisms between fiber bundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.

\section{Conclusion}
The data model and functional transform refactor presented here allows bulding block libraries to better support domain specific libraries without having to explicitly take in the specific data structure and visualization needs of those domains. Adopting this model would induce a separation of data representation and visual representation that, for example, in Matplotlib is so entangled that it has lead to a brittle and sometimes incoherent API and internal code base. A refactor driven by TEAM would result in components that can be be guaranteed to preserve continuity and structure, as defined by the domain specific library developers, without having to fold domain specific assumptions back into the base library.

%% if specified like this the section will be committed in review mode
\acknowledgments{
The authors wish to thank the Matplotlib development team for their invaluable feedback along the way, particulary Bruno Beltran, Eric Schles and Chana Tilevitz for idenfiying the confusing bits, Nicolas Kruchten for articulating the framing and Marc Hanwell, Lev Manovich, Robert Haralick and Huy Vo for invaluable feedback on various iterations. This paper is indebted to Kindlmann and Scheidegger's \cite{kindlmannAlgebraicProcessVisualization2014} for serving as a model and Munzner's guide to writing visualization papers. This project has been made possible in part by grant number 2019-207333 from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation.}

>>>>

%% declare the category of your paper, only shown in review mode
\vgtccategory{Research}
%% please declare the paper type of your paper to help reviewers, only shown in review mode
%% choices:
%% * algorithm/technique
%% * application/design study
%% * evaluation
%% * system
%% * theory/model
\vgtcpapertype{theory/model}

@@

General purpose visualization libraries are often tasked with supporting many domains, leading the libraries to grow organically in ways that can lead to incoherent interfaces and brittle components. To adapt to modern library development needs, this paper presents a functional model of the how components provided by lower level libraries can be guaranteed to be structure preserving maps from data to graphic. The target for this model are the libraries developers use to build automated and user facing tools rather than those tools. The model presented here is agnostic towards what makes for a good visualization, so long as the specified constraints for building those visualizations are satisfied. 
\begin{figure}[H]
  \centering 
  \subfloat[\label{fig:intro:continuity}]{%
    \includegraphics[width=1\columnwidth]{continuity.png}}\\
  \subfloat[\label{fig:intro:equivariance}]{%
    \includegraphics[width=1\columnwidth]{equivariance.png}}
\caption{In \autoref{fig:intro:continuity}, the line plot does not preserve \textit{continuity} because it implies that the discrete records are connected to each other, while the bar plot is \textit{continuity} preserving because it visually represents the records as independent data points. In \autoref{fig:intro:equivariance}, the data in blue is scaled by a factor of two, yielding the data in orange. To preserve \textit{equivariance}, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
\label{fig:intro:structure}
\end{figure}
The structure that components of a visualization library must preserve is \textit{continuity} and \textit{equivariance}, as illustrated in \autoref{fig:intro:structure}. \textit{Continuity} is the way in which the records in the dataset are connected to each other and \textit{equivariance} means that if any action is applied to the data or the visual, for example a rotation, permutation, translation, or rescaling, an equivalent action is applied on the other side. In \autoref{fig:intro:continuity}, the line plot does not preserve continuity because the line connecting the discrete categories implies that the frequency of weather events is sampled from a continuous interval and the categories are points on that interval. In \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from blue to orange data and lines.

To specify the constraints components must satisfy to be equivariant continuity preserving mappings from data to visual representation, we developed a functional model, which we call the Topological Equivariant Artist Model (TEAM). Motivated by the challenge of re-architecturing the Python visualization library Matplotlib, a functional architecture derived from our model would gain a generalized data abstraction and well specified reusable components with minimal side effects\cite{loudenProgrammingLanguagesPrinciples2010}, in turn yielding improvements in maintainability, extendability, scaling, and support for concurrency. The contribution of this work is 
\begin{enumerate}
  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:math:graphic:base}
  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:math:artist}
  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:code}
\end{enumerate}

\section{Related Work}
The notion that visualization is equivariant continuity preserving maps from data to visual representation is neither a new formalism nor a new implementation goal; this work bridges the formalism and implementation in a functional manner with a topological approach at a building blocks library\cite{wongsuphasawatNavigatingWideWorld2021} level. When introducing the retinal variables, Bertin informally specifies that continuity is preserved in the mark and defines equivariance constraints in terms of data and visual variables being selective, associative, ordered, or quantitative\cite{bertinSemiologyGraphicsDiagrams2011a}. In the \textit{A Presentation Tool}(APT) model, Mackinlay embeds the continuity constraint in the choice of visualization type and generalizes the equivariance constraint to preserving a binary operator from one domain to another. The algebraic model of visualization\cite{kindlmannAlgebraicProcessVisualization2014}, proposed by Kindlmann and Scheidegger, restricts equivariance to invertible transformations. Our model defines equivariance in terms of monoid actions, which are more restrictive than all binary operations but more general than invertible transformations. The Kindlmann and Scheidegger algebraic model also describes visualization as commutative mappings from data space to representation space to graphic space. Our model has the same notion of commutative mappings across analogous spaces, but uses topology to explicitly include continuity.

We emphasize continuity because visualization tools tend to either be architectured around a core data structure\cite{HeerSoftware2006} or explicitly implement the visual algorithm in terms of the data continuity the algorithm expects \cite{toryRethinkingVisualizationHighlevel2004}. For example, the relational database is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}. Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}. Neither the table nor image nor graph model on its own supports all the data types a typical general purpose visualization library needs to support; instead libraries such as Matplotlib\cite{hunterMatplotlib2DGraphics2007} and Vtk\cite{hanwellVisualizationToolkitVTK2015, geveci2012vtk} and D3 \cite{bostockDataDrivenDocuments2011} explicitly carry around different data representations for all the different types of visualizations they support.  Where libraries with a single core data structure have very consistent APIs, VTK, D3 and Matplotlib APIs can seem incoherent as every visualization has a different notion of how the data is structured. Our model facilitates unifying these APIs via an abstraction of data general enough to encompass most continuities, the clear separation of data, visual, and graphic transformations into functional components to mitigate side effects, and by specifying the constraints these components must satisfy to map data to visualizations in a structure preserving manner.

@@

In \autoref{fig:teaser}, we illustrate the output of a minimal \vmark\ that will generate distinguishable graphical marks: non-overlapping scatter points, a non-infinitely thin line, and an image. The scatter plot can be defined as 
\begin{equation}
\vmark(xpos, ypos)(\alpha, \beta)
\label{eq:math:artist:q:scatter}
\end{equation}
with a constant \textit{size} and color $\gsection_{RGB} = (0,0,0)$ that are defined as part of \vmark. The position of this swatch of color can be computed relative to the location on the disc \((\alpha, \beta) \in \gbase_{\dbasepoint}\) as shown in \autoref{fig:math:artist:base}
\begin{align*}
x &= size *\alpha \cos(\beta) + xpos \\
y &= size *\alpha \sin(\beta) + ypos
\end{align*}
such that $\gsection(\gbasepoint) = (x, y, 0, 0, 0)$ colors the point (x,y) black. 
In contrast, the line plot 
\begin{equation}
\vmark(xpos, \hat{n}_{1}, ypos, \hat{n}_{2})(\alpha, \beta)
\label{eq:math:artist:q:line}
\end{equation}
in \autoref{fig:teaser} has a \vindex\ function that is not only parameterized on \dbasepoint\ but also on the $\alpha$ distance along \dbasepoint\ and corresponding region in \gbase. As shown in \autoref{fig:math:artist:base}, line needs to know the tangent of the data to draw an envelope above and below each (xpos,ypos) such that the line appears to have a thickness; therefore the artist takes as input the jet bundle \cite{JetBundle2020,musilovaCalculusVariationsJet2016} $\mathcal{J}^{2}(\dtotal)$ which is the data \dtotal\ and the first and second derivatives of \dtotal. The magnitude of the slope is $\lvert n \rvert = \sqrt{{n_{1}}^2 + {n_{2}}^2}$
such that the normal is  $\hat{n}_{1} = \frac{n_1}{\lvert n \rvert}, \; \hat{n}_{2} = \frac{n_2}{\lvert n \rvert}$ which yields components of \gsection\
\begin{align*}
 x = xpos(\vindex(\alpha)) &+ width*\beta\hat{n}_1(\vindex(\alpha)) \\
 y = ypos(\vindex(\alpha)) &+ width*\beta\hat{n}_2(\vindex(\alpha)) 
\end{align*}
where (x,y) look up the position $\vindex(\alpha)$ on the data and the derivatives $\hat{n}_1, \hat{n}_2$ . The derivatives are then multiplied by a $width$ parameter to specify the thickness.
In \autoref{fig:teaser}, the image 
\begin{equation}
\vmark(xpos, ypos, color)
\label{eq:math:artist:q:image}
\end{equation}
 is a direct lookup into  $\vindex:\gbase\rightarrow\dbase$. The indexing variables $(\alpha, \beta)$ define the distance along the space, which is then used by \vindex\ to map into \dbase\ to lookup the color values 
\begin{equation*}
R = R(\vindex(\alpha, \beta)), \; G = G(\vindex(\alpha, \beta)), \; B = B(\vindex(\alpha, \beta))
\end{equation*}
In the case of an image, the indexing mapper \vindex\ may do some translating to a convention expected by \vmark, for example reorientng the array such that the first row in the data is at the bottom of the graphic.

@@

\subsubsection{Visual Encoders}
The visual parameter serves as the dictionary key because the visual representation is constructed from the encoding applied to the data  $\vsection = \vchannel \circ \dsection$. For the scatter plot, the mappings for the visual fiber components $\vfiber=(x,y, facecolors, s)$ are defined as
\begin{minted}{python}
cmap =  color.Categorical({'true':'deeppink', 
                           'false':'deepskyblue'}) 
# {P_i name:{'name':c_i, 'encoder:\nu_i}}
V = {'x': {'name': 'v4', 'encoder': lambda x: x}, 
     'y': {'name': 'v2', 'encoder': lambda x: x},
     'facecolors': {'name':'v3', 'encoder': cmap}, 
     's':{'name': None , 
          'encoder': lambda _: itertools.repeat(.02)}}
\end{minted}
where \mintinline{python}{lambda x: x} is an identity \vchannel, \mintinline{python}|{'name':None}| maps into \vfiber\ without corresponding \dsection\ to set a constant visual value, and \mintinline{python}{color.Categorical} is a custom \vchannel\ implemented as a class for reusability. 
\begin{minted}{python}
#\nu_i(m_r(E_i)) = \varphi(m_r)(\nu_i((E_i))
def test_nominal(values, encoder):
    m1 = list(zip(values, encoder(values)))
    random.shuffle(values)
    m2 = list(zip(values, encoder(values)))
    assert sorted(m1) == sorted(m2)
\end{minted}
As described in \autoref{eq:math:artist:nu:equivariance}, a test for equivariance can be implemented trivially. It is currently factored out of the artist for clarity.

@@

\section{Discussion}
This work contributes a functional model of the structure-preserving maps from data to visual representation to guide the development of visualization libraries, thereby providing a means to express the constraints of preserving the data continuity in the graphic and faithfully translating the properties of the data variables into visual variables. Combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets this model support a variety of datasets, including  discrete relational tables, multivariate high resolution spatio temporal datasets, and complex networks. Decomposing the artist into encoding \vchannel, assembly \vmark, and reindexing \vindex\ provides the specifications that the graphic must have continuity equivalent to the data, and that the visual characteristics of the graphics are equivariant to their corresponding components under monoid actions. This model defines these constraints on the transformation function such that they are not specific to any one type of encoding or visual characteristic. Encoding the graphic space as a fiber bundle provides a structure rich abstraction of the target graphical design in the target display space. The toy prototype built using this model validates that is usable for a general purpose visualization tool since it can be iteratively integrated into the existing architecture rather than starting from scratch. Factoring out graphic formation into assembly functions allows for much more clarity in how they differ. This prototype demonstrates that this framework can generate the fundamental point (scatter plot) and line (line chart) marks.

@@

\subsection{Future Work}
More work is needed to formalize the composition operators and equivalence class \vartisteq. We also need to implement artists that demonstrate that the model can underpin a minimally viable library, foremost an image\cite{haber1990visualization,hansen2011visualization}, a heatmap\cite{wilkinsonHistoryClusterHeat2009,loua1873atlas}, and an inherently computational artist such as a boxplot\cite{wickham40YearsBoxplots2011}. Since this model formalizes notions of structure preservation, it can serve as a good base for tools that assess quality metrics\cite{bertiniQualityMetricsHighdimensional2011a} or invariance \cite{kindlmannAlgebraicProcessVisualization2014}. While this paper formulates visualization in terms of monoidal action homomorphisms between fiber bundles, the model lends itself to a categorical formulation\cite{fongInvitationAppliedCategory2019,milewskiCategoryTheoryProgrammers} that could be further explored.

\section{Conclusion}
The data model and functional transform refactor presented here allows building block libraries to better support downstream, including domain specific, libraries without having to explicitly incorporate the 
specific data structure and visualization needs of those domains without having to fold domain specific assumptions back into the base library. Adopting this model would induce a separation of data representation and visual representation that, for example, in Matplotlib is so entangled that it has lead to a brittle and sometimes incoherent API and internal code base. A TEAM driven refactor would produce more maintainable, reusable, and extensible code, leading to better building blocks for the ecosystem of tools built on top of Matplotlib.

%% if specified like this the section will be committed in review mode
\acknowledgments{
The authors wish to thank the Matplotlib development team for their invaluable feedback along the way, particulary Bruno Beltran, Eric Schles and Chana Tilevitz for idenfiying the confusing bits, Nicolas Kruchten for articulating the framing, and Marc Hanwell, Lev Manovich, Robert Haralick and Huy Vo for very helpful feedback on various iterations. This paper is indebted to Kindlmann and Scheidegger's \cite{kindlmannAlgebraicProcessVisualization2014} for serving as a model and Munzner's guide to writing visualization papers. This project has been made possible in part by grant number 2019-207333 from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation.}
